---
title: "Spring 2025 Project"
author: "Armelle Duston"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    extra_dependencies:
      - bm
      - amsmath
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(warning = F, message = F, echo = F)

if (rstudioapi::isAvailable()) {
  setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
}

packages <- c("ggplot2", "dplyr", "httr", "tidycensus", 
              "stringr", "spdep", "kableExtra", "viridis", "RColorBrewer")
invisible(lapply(packages, library, character.only = TRUE))

# Install INLA separately since it's not on CRAN
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if ("INLA" %in% rownames(installed.packages())==FALSE){
  install.packages(
  "INLA", 
  repos = c(
    INLA = "https://inla.r-inla-download.org/R/stable",
    CRAN = "https://cloud.r-project.org"
  ),
  dependencies = TRUE
)
}
library(INLA)

```


# Data

See data_dict.txt for more information on variables in the dataset. Briefly, the outcome variable of interest is lung cancer incidence rate (age-adjusted to the 2000 US standard population, per 100,000) during the years 2010-2014. The primary covariate of interest in average PM2.5 (micrograms/meter^3) from 2000-2005. Other covariates considered are EPA's environmental quality indexes (see EPA EQI report) from the same time period. 

```{r}

dataverse_server <- "https://dataverse.harvard.edu"
file_id <- ":persistentId?persistentId=doi:10.7910/DVN/HMOEJO/D3GVNB" 
api_key <- "8c1d730e-50a2-426d-a6ce-e65e8ba5318f"
endpoint <- paste0(dataverse_server, "/api/access/datafile/", file_id)

response <- GET(endpoint, add_headers("X-Dataverse-key" = api_key))

if (status_code(response) == 200) {
  
  data_text <- content(response, as = "text", encoding = "UTF-8")
  my_data <- read.delim(textConnection(data_text), 
                        header = TRUE, 
                        sep = "",
                        stringsAsFactors = FALSE)
  
} else {
  cat("Request failed with status code:", status_code(response), "\n")
}

```


```{r}

# Get state and county shapefiles and population counts from tidycensus and join with main df

# set Census API key 
# census_api_key("491a326a4f51b9f77b828740ce4f0571f2951d8d", 
#                install = TRUE, overwrite = FALSE)
# readRenviron("~/.Renviron")

# get county geometries + population for 2010
counties_data <- get_acs(
  geography = "county",
  variables = "B01003_001",       # total population
  year      = 2010,
  geometry  = TRUE
)
counties <- counties_data |>
  rename(population2010 = estimate) |>
  select(GEOID, geometry, population2010)


# join with main data
my_data <- my_data |> 
  mutate(GEOID = str_pad(
      as.character(FIPScode),
      width = 5,
      side  = "left",
      pad   = "0"
    )) |>
  left_join(counties, by = "GEOID")

# get state shapefiles 
states <- get_acs(geography = "state", 
                  variables = "B01003_001", 
                  year = 2010, 
                  geometry = TRUE) |>
    st_as_sf() |>
    filter(!NAME %in% c("Alaska", "Hawaii", "Puerto Rico"))

```

\newpage

# Modeling

I considered two models for the lung cancer incidence per 100,000 (for the time period 2010-2014): normal and gamma. The incidence is positive and continuous, so a gamma model could be appropriate particularly if it is right-skewed. To model spatial variation, I used a Bayesian spatial model as follows (Besag-York-Mollie (BYM) model):


$$
\begin{aligned}
  Y_i &\sim Gamma(\alpha, \frac{\mu_i}{\alpha})  \\
  \log(\mu_i) &= z_i \beta + u_i + v_i \\ \\
  Y_i &\sim Normal(\mu_i, \sigma^2) \\
  \mu_i &= z_i \beta + u_i + v_i \\ \\
  z_i: &\text{ vector of covariates (fixed)} \\
  u_i: &\text{ spatial random effect} \\
  v_i: &\text{ uncorrelated noise} \\ \\
  u_i|u_{-i} &\sim Normal(\bar{u}_{\delta_i}, \frac{\sigma^2_u}{n_{\delta_i}}), (\delta_i \text{ is the set of neighbors}) \\
  v_i &\sim Normal(0, \sigma_v^2)
\end{aligned}
$$


In particular, I am interested in What is the association between PM2.5 and rate of lung cancer mortality? The following covariates are considered for the models:

- Primary interest
  -PM2.5
    
- Possible confounders
  - LandEQI
  - SociodemographicEQI
  - BuiltEQI

\newpage

# Exploratory analysis

## Distribution of variables of interest

```{r}

my_data |> ggplot(aes(x=LungCancer)) +
  geom_histogram() +
  labs(title = "Distribution of lung cancer incidence") + 
  theme_minimal()

my_data |> ggplot(aes(x=PM2.5)) +
  geom_histogram() +
  labs(title = "Distribution of PM2.5") + 
  theme_minimal()
  
  
```

\newpage

## Maps of variables of interest

```{r, fig.width=6, fig.height=4, fig.show='hold'}

# Data for mapping - make sf obj and filter alaska and hawaii
my_data_sf <- st_as_sf(my_data) |>
  filter(!str_starts(GEOID, "02") & !str_starts(GEOID, "15"))

# Function to generate maps
create_map <- function(data, var_name, title, fill_color = "viridis") {
  
  # Handle NA values in the scale
  scale_fill <- if(fill_color == "viridis") {
    scale_fill_viridis_c(na.value = "grey80", 
                         option = "plasma", 
                         name = var_name)
  } else {
    scale_fill_distiller(na.value = "grey80", 
                         palette = fill_color, 
                         name = var_name, 
                         direction = 1)
  }
  
  ggplot() +
    geom_sf(data = my_data_sf, 
            aes(fill = .data[[var_name]], geometry = geometry), 
            color = NA, size = 0) +
    geom_sf(data = states, fill = NA, color = "black", size = 0.1) +
    scale_fill +
    coord_sf(datum = NA) +
    theme_minimal() +
    labs(title = title) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      panel.background = element_rect(fill = "white", color = NA)
    )
}

# Create maps for each variable
pm25_map <- create_map(my_data_sf, 
                       "PM2.5", 
                       "PM2.5 Concentration by County")

aireqi_map <- create_map(my_data_sf, 
                         "AirEQI", 
                         "Air Environmental Quality Index by County")

landeqi_map <- create_map(my_data_sf, 
                          "LandEQI", 
                          "Land Environmental Quality Index by County")

sociod_map <- create_map(my_data_sf, 
                         "SociodEQI", 
                         "Sociodemographic Environmental Quality Index by County")

built_map <- create_map(my_data_sf, 
                         "BuiltEQI", 
                         "Built Environmental Quality Index by County")

lung_cancer_map <- create_map(my_data_sf, 
                              "LungCancer", 
                              "Lung Cancer Rate by County", 
                              "YlOrRd")

# Display the maps
pm25_map
aireqi_map
landeqi_map
sociod_map
built_map
lung_cancer_map

```

# Correlation plot

```{r}

variables_of_interest <- c("PM2.5", "AirEQI", "LandEQI", "SociodEQI", "BuiltEQI", "LungCancer")
cor_data <- my_data |> 
  select(all_of(variables_of_interest)) |>
  cor(use = "pairwise.complete.obs")
corrplot::corrplot(cor_data, method = "circle", type = "upper", 
                   tl.col = "black", tl.srt = 45)

```

\newpage

# Building & fitting model

Focus modeling on eastern states: they have areas with lower and higher incidence of lung cancer, lower missingness of data, and smaller counties (finer resolution). 

```{r}

# keep eastern states
reduced_data <- my_data_sf |>
  filter(State %in% c("VA", "WV", "KY", "TN", "NC", "SC", "GA", 
                      "FL", "AL", "MS", "LA", "MO", "AR",
                      "IA", "WI", "IL", "IN", "OH", "PA",
                      "MD", "DL", "DC", "NJ", "NY", "CT",
                      "RI", "MA", "VT", "NH", "ME"))
  
  
# neighborhood matrix 
neighbors <- poly2nb(reduced_data)
nb2INLA("reduced_data.adj", neighbors)
graph <- inla.read.graph(filename = "reduced_data.adj")

reduced_data$re_u <- 1:nrow(reduced_data)
reduced_data$re_v <- 1:nrow(reduced_data)

fit_spatial_model <- function(covariates){
  
  formula <- paste("LungCancer ~ ",
                   covariates, 
                   "+ f(re_u, model = 'besag', graph = graph, scale.model = TRUE) +", 
                   # spatial random component 
                   "f(re_v, model = 'iid')" # unstructured random component
                   )


  gamma_model <- inla(as.formula(formula), family = "gamma", 
                      data = reduced_data,
                      control.predictor = list(compute = TRUE),
                      control.compute = list(return.marginals.predictor = TRUE, 
                                             waic = TRUE))
  
  normal_model <- inla(as.formula(formula), family = "gaussian", 
                      data = reduced_data,
                      control.predictor = list(compute = TRUE),
                      control.compute = list(return.marginals.predictor = TRUE, 
                                             waic = TRUE))


  return(list(gamma_model, normal_model)) 
  
}

cov1 <- "PM2.5"
cov2 <- "PM2.5 + LandEQI + SociodEQI + BuiltEQI"
cov3 <- "AirEQI"
cov4 <- "AirEQI + LandEQI + SociodEQI + BuiltEQI"


model1 <- fit_spatial_model(cov1)
model2 <- fit_spatial_model(cov2)
# model3 <- fit_spatial_model(cov3)
# model4 <- fit_spatial_model(cov4)
  
```

Check out residuals (look for residual spatial clustering)

```{r, fig.width=6, fig.height=4, fig.show='hold'}

east_states <- states |>
  mutate(centroid = st_centroid(geometry)) |>
  mutate(lon = st_coordinates(centroid)[,1]) |>
  filter(lon > -94) |>
  select(-centroid, -lon)


plot_resid <- function(fit) {
  resid_df <- reduced_data |> select(geometry)
  resid_df$mu   <- fit$summary.fitted.values$mean
  resid_df$pear <- (reduced_data$LungCancer - resid_df$mu) / sqrt(resid_df$mu)
  
  p1 <- ggplot(resid_df) +
    geom_sf(aes(fill = pear), color = NA, size = 0) +
    geom_sf(data = east_states, fill = NA, color = "black", size = 0.1) +
    scale_fill_distiller(
      palette  = "RdBu",
      direction = -1,
      limits   = c(-max(abs(resid_df$pear)), max(abs(resid_df$pear))),  
      name     = "Pearson\nresidual") +
    coord_sf(datum = NA) +
    theme_minimal() +
    labs(title = "Pearson Residuals") +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      panel.background = element_rect(fill = "white", color = NA)
    )
  return (p1)
}

plot_resid(model1[[1]])
plot_resid(model2[[1]])
plot_resid(model1[[2]])
plot_resid(model2[[2]])


```

Gamma models (first two plots) appear to adequately capture the spatial variation (residuals look random), normal models (second two plots) still have some hotspots/clustering 

\newpage

Results for models with PM2.5

```{r}

### extract summaries from models and adjust as below
get_sum <- function(fit) {
  summary <- fit$summary.fixed |>
    rename(lower = `0.025quant`, upper = `0.975quant`)
}

results_df1 <- data.frame(rbind(exp(get_sum(model1[[1]])),
                                exp(get_sum(model2[[1]])),
                                get_sum(model1[[2]]),
                                get_sum(model2[[2]])
                                )
                          ) %>%
  filter(!grepl("Intercept", rownames(.))) |>
  mutate(mean_ci = sprintf(
      "%.2f (%.3f, %.3f)",  
      mean, lower, upper) ) |>
  select(mean_ci)

get_waic <- function(fit) {
  return ( as.character(round(fit$waic$waic), digits=0))
}


results_df1 <- cbind(results_df1[1:5,], 
                     rbind(get_waic(model1[[1]]),
                           get_waic(model2[[1]]),
                           cbind(rep(NA, 3))),
                     results_df1[6:10,],
                     rbind(get_waic(model1[[2]]),
                           get_waic(model2[[2]]),
                           cbind(rep(NA, 3)))) 


rownames(results_df1) <- c("PM2.5 (unadjusted)",
                           "PM2.5 (adjusted)",
                           "LandEQI",
                           "SociodEQI",
                           "BuiltEQI")
colnames(results_df1) <- c("(multiplicative)", "WAIC-gam", "(additive)", "WAIC-gaus")
  
results_df1 |> 
  kbl(caption = "Lung Cancer incidence models ", 
      booktabs = TRUE) |>
    kable_styling(full_width = FALSE,
                  latex_options = c("hold_position", "striped")) |>
  add_header_above(c(" " = 1,
                     "Gamma model" = 2, 
                     "Gaussian model" = 2)) |>
  group_rows("Unadjusted", 1, 1) |>
  group_rows("Adjusted", 2, 5) 


```

Gamma model indicates no effect of PM2.5 (both sets of covariates) and modest effect of landEQI and sociodEQI. Gaussian model indicates modest effect of PM2.5 (both sets of covariates). Effect of landEQI goes in opposite direction than in gamma model, sociodEQI and builtEQI also significant.


\newpage

# Appendix: R Code
```{r getlabels, echo = FALSE}
labs = knitr::all_labels()
labs = labs[!labs %in% c('setup', 'getlabels')] ### Exclude code from this chunk and the set up chunk
```

```{r allcode, ref.label = labs, eval = FALSE, echo = T}
```









